---
title: "R Notebook"
output: github_document
---

# CRMetrics class


## Initializing a CRMetrics class

Source the file in which the class is defined. 

```{r}
source("create_R6_class.R")
```

Initialize a new object of class `CRMetrics` with the path to the Cell Ranger output and a metadata file. 
The metadata file contains a column `sample` with the samples and optionally more columns with factors. 

```{r}
crmetrics <- CRMetrics$new(data_path = "/data/PD-MSA_lentiform_nucleus/counts_premrna/", 
                           metadata_file = "../data/metadata.csv")
```

We populated the metadata and the summary metrics field of the object when initializing it. 

```{r}
head(crmetrics$metadata)
head(crmetrics$summary_metrics)
```


## Plotting 

Now we can already create plots from the summary statistics. 
E.g., `Median UMI Counts per Cell` is defined in the Cell Ranger summary statistics of each sample. 

The comparison group is a column in the metadata file. 
It will be the variable on the x-axis and pairwise statistical comparisons will be made on these groups. 

```{r}
crmetrics$plot_median_umi(comp_group = "sex")
crmetrics$plot_median_umi(comp_group = "group")
```

If no comparison group is specified, the samples are plotted on the x-axis.  
This is potentially not very meaningful but it ensures that the plots work even if we don't have groups in the samples. 

```{r}
crmetrics$plot_median_umi()
```


Specifying comparison that is not a column in the metadata will throw an error.

```{r error=TRUE}
crmetrics$add_comparison("gropu")
```

We can add a comparison group globally which will be the default if not specified in the plotting function

```{r}
crmetrics$add_comparison("group")
crmetrics$plot_median_umi()
```

A possibility to reset the comparison group is to se the field in the class to `NULL`. 

```{r}
crmetrics$comp_group <- NULL
```


We can also create plots on statistics of the count matrices.  
This requires to load the detailed metrics and will some time. 

```{r}
crmetrics$add_detailed_metrics()


# this is slow so we can also cheat if we already have the file

# metadata <- read.csv("../data/metadata.csv")
# detailed_metrics <- read_detailed_metrics(samples = metadata$sample, data_path = "/data/PD-MSA_lentiform_nucleus/counts_premrna/")

# crmetrics$detailed_metrics <- detailed_metrics
```

We can plot the distribution of number of UMIs and expressed genes in each sample. 

```{r}
crmetrics$plot_gene_count()
crmetrics$plot_umi_count()
```

And of course add group information. 

```{r}
crmetrics$plot_gene_count(comp_group = "group")
```
